---
title: "BIMM 143 Class 17"
author: Xaler Lu (A17388454)
toc: true
format: pdf
---

# Unix Work

## Background

We will use AWS EC2 instances to analyze large biomolecular sequencing datasets. 

Don't forget the IP address of the cloud computer changes every time you relaunch an instance. 

`ssh -i "~/Downloads/BIMM143_Class16_xaler.pem" ubuntu@ec2-35-87-198-206.us-west-2.compute.amazonaws.com`

## The Sequence Read Archive (SRA)

We will visit NCBI's major repository, SRA, for sequencing data. The files we will obtain are highly-compressed, so we must use special tools to decompress these records. 

To begin, we will install the **SRA-toolkit** software and unzip it.

`curl -O https://ftp-trace.ncbi.nlm.nih.gov/sra/sdk/current/sratoolkit.current-ubuntu64.tar.gz  `

`tar -zxvf sratoolkit.current-ubuntu64.tar.gz`

We will then go into the directory. Now, we don't want to type out the entire `sratoolkit.3.3.0-ubuntu64/bin/prefetch` every time we want to run `prefetch`, so use the following.

`export PATH=$PATH:/home/ubuntu/sratoolkit.3.3.0-ubuntu64/bin`

## ChIP-Seq Data from Ovarian Cancer DNA Sample

<https://www.ncbi.nlm.nih.gov/sra/SRX193578>

Use `prefetch` to run it with the SRR ID we found in the link.

Prepare the temporary files \
`prefetch SRR600956`

Exract the data in FASTQ format in your current directory \
`fasterq-dump SRR600956`

## RNA-Seq Data

The paper <https://pubmed.ncbi.nlm.nih.gov/26656844/> \
The GEO entry <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE72001>

We will look into an RNA-Seq dataset from a human cancer cell line.

We will start with the SRA accession number, download their raw data, and analyze the reads for PCA and differential expression analysis.

Prepare the temporary files \
`prefetch SRR2156848`

Exract the data in FASTQ format in your current directory \
`fasterq-dump SRR2156848`

Now let's download three other datasets. \
`prefetch SRR2156849 SRR2156850 SRR2156850` \
`fastq-dump SRR2156849 SRR2156850 SRR2156850`

## Transcript Quantification via Pseudoalignment

**Kallisto** compares each RNA-seq read against an indexted table of transcript sequences 

`wget https://github.com/pachterlab/kallisto/releases/download/v0.44.0/kallisto_linux-v0.44.0.tar.gz`

We will use the Ensembl reference transcriptome

`wget ftp://ftp.ensembl.org/pub/release-67/fasta/homo_sapiens/cdna/Homo_sapiens.GRCh37.67.cdna.all.fa.gz`

`gunzip Homo_sapiens.GRCh37.67.cdna.all.fa.gz`


The `index` command builds the transcript index \
`kallisto index -i hg19.ensembl Homo_sapiens.GRCh37.67.cdna.all.fa`

The `quant` commands quantifies each transcripts \
`kallisto quant -i hg19.ensembl -o SRR2156848_quant SRR2156848_1.fastq SRR2156848_2.fastq`

`kallisto quant -i hg19.ensembl -o SRR2156849_quant SRR2156849_1.fastq SRR2156849_2.fastq`

`kallisto quant -i hg19.ensembl -o SRR2156850_quant SRR2156850_1.fastq SRR2156850_2.fastq`

`kallisto quant -i hg19.ensembl -o SRR2156851_quant SRR2156851_1.fastq SRR2156851_2.fastq`

When we have repetitive scripting tasks, we can create simple functions

`nano run_kallisto.sh`

In the nano function, imput the following code. 
> for id in SRR2156849 SRR2156850 SRR2156851; do
  echo "Processing sample: $id"
  kallisto quant -i hg19.ensembl -o ${id}_quant ${id}_1.fastq ${id}_2.fastq
done

echo "All samples complete!"

## Transfering the results

With these new files, we want to transfer them from the AWS instance to our computer. To do this, open a new terminal and input the following.

`scp -r -i "~/Downloads/BIMM143_Class16_xaler.pem" ubuntu@ec2-35-87-198-206.us-west-2.compute.amazonaws.com:~/*_quant .`

# Downstream Analysis

We will finish the analysis in R with the `tximport` package.
```{r, message=FALSE, warning=FALSE}
library(tximport)
library(rhdf5)
```


We will then set up the folders and file names
```{r}
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path(folders, "abundance.h5")
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

We now have our estimated transcript counts for each sample
```{r}
colSums(txi.kallisto$counts)
```

```{r}
sum(rowSums(txi.kallisto$counts) >0)
```

We want to also filter out those annotated transcripts with no reads

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

We then filter out those with no changes over the samples
```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## PCA

```{r}
pca <- prcomp(t(x), scale=TRUE)
summary(pca)
```
Base R PCA Plot
```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

PCA ggplot
```{r, message=FALSE,warning=FALSE}
library(ggplot2)
library(ggrepel)
```

```{r}
# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)
```

### PC1 VS PC2
```{r}
ggplot(y) +
  aes(PC1, PC2, col = Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

### PC1 VS PC3
```{r}
ggplot(y) +
  aes(PC1, PC3, col = Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

### PC2 VS PC3
```{r}
ggplot(y) +
  aes(PC2, PC3, col = Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```