---
title: "BIMM 143 Class 18"
author: Xaler Lu (A17388454)
toc: true
format: pdf
---

## Background

Whooping cough is a highly contagious respiratory disease caused by the bacterium *Bordetella pertussis*. The toxin released destroys the cilia in the lungs. It develops in three main phases, and has two vaccine: whole-cell and acellular. Whole-cell vaccine has the whole bacterial cell with the endotoxin. Acellular has the most important components that the body recognizes and has less side effects. 

## Investigating Pertussis

The CDC has compiled data since 1922 found, but there is no way to download the CSV file. Thus, we will scrap the website for the data using the `datapasta` package. Then, the `Addins` tab should appear at the top of RStudio. We will copy the table at the CDC website, go to `Addins`, and paste as data frame.

```{r, message=FALSE,warning=FALSE}
library(ggplot2)
library(datapasta)
```

```{r, echo = FALSE}
cdc <- data.frame(Year = c(1922L,1923L,1924L,1925L,
                           1926L,1927L,1928L,1929L,1930L,1931L,
                           1932L,1933L,1934L,1935L,1936L,
                           1937L,1938L,1939L,1940L,1941L,1942L,
                           1943L,1944L,1945L,1946L,1947L,
                           1948L,1949L,1950L,1951L,1952L,
                           1953L,1954L,1955L,1956L,1957L,1958L,
                           1959L,1960L,1961L,1962L,1963L,
                           1964L,1965L,1966L,1967L,1968L,1969L,
                           1970L,1971L,1972L,1973L,1974L,
                           1975L,1976L,1977L,1978L,1979L,1980L,
                           1981L,1982L,1983L,1984L,1985L,
                           1986L,1987L,1988L,1989L,1990L,
                           1991L,1992L,1993L,1994L,1995L,1996L,
                           1997L,1998L,1999L,2000L,2001L,
                           2002L,2003L,2004L,2005L,2006L,2007L,
                           2008L,2009L,2010L,2011L,2012L,
                           2013L,2014L,2015L,2016L,2017L,2018L,
                           2019L,2020L,2021L,2022L,2023L,2024L,                                 2025L),
                 Cases = c(107473,164191,165418,152003,
                           202210,181411,161799,197371,
                           166914,172559,215343,179135,265269,
                           180518,147237,214652,227319,103188,
                           183866,222202,191383,191890,109873,
                           133792,109860,156517,74715,69479,
                           120718,68687,45030,37129,60886,
                           62786,31732,28295,32148,40005,
                           14809,11468,17749,17135,13005,6799,
                           7717,9718,4810,3285,4249,3036,
                           3287,1759,2402,1738,1010,2177,2063,
                           1623,1730,1248,1895,2463,2276,
                           3589,4195,2823,3450,4157,4570,
                           2719,4083,6586,4617,5137,7796,6564,
                           7405,7298,7867,7580,9771,11647,
                           25827,25616,15632,10454,13278,
                           16858,27550,18719,48277,28639,32971,
                           20762,17972,18975,15609,18617,
                           6124,2116,3044,7063,22538,21996)
       )
```


```{r}
head(cdc)
```

> Q. Plot of year vs cases

```{r}
ggplot(cdc) +
  aes(Year, Cases) +
  geom_point() +
  geom_line(col = "purple") +
  labs(title = "Cases per Year") +
  theme_bw()
```

> Q. Add annotation lines for the major milestones of the wP vaccination roll-out (1946) and switch to the aP vaccine (1996).

```{r}
ggplot(cdc) +
  aes(Year, Cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col ="blue", lty = 2) +
  geom_vline(xintercept = 1996, col = "purple", lty = 2) +
  geom_vline(xintercept = 2020, col = "grey", lty = 2) +
  labs(title = "Cases per Year") +
  theme_bw()
```

> Q2. What do you notice when wP and aP vaccines were introduced?

Number of cases dropped when wP was introduced, but not much of an effect for aP. This showed the vaccines was highly effective in reducing infection.

> Q3. What is the explanation for after aP vaccine was introduced?

The aP vaccination roll-out didn't have a strong effect or less people got vaccinated. Maybe the methods of measuring cases got better, so the overall cases went up. 

## Exploring CMI-PB

Why is this vaccine-preventable disease on the upswing? We will need to investigate the mechanisms underlying waning protection against pertussis. The CMI-PB project tracks long-term humoral and cellular immune response to people who received the wP or aP vaccines followed by aP booster vaccines.

The data is in JSON format that uses a series of key-value pairs. <https://www.cmi-pb.org> We read the JSON file using `read_json()`

```{r, message=FALSE,warning=FALSE}
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)

head(subject,3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

There are 87 aP vaccinated and 85 wP vaccinated individuals.

> Q5. How many male and female subjects are in the dataset?

```{r}
table(subject$biological_sex)
```

There are 112 females and 60 males in the dataset. 

> Q6. What is the breakdown of race and biological sex?

```{r}
table(subject$race, subject$biological_sex)
```
> Q. Is this reprsentative of the US population? 

This does not represent the US population because this research was done during the COVID pandemic. 

We can read more tables in the CMI-PB database
```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = T)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = T)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```
We want to know the subject's general information, vaccinations, and antibody titer. 
Let's merge these data frames together using `link()` or `*_join()` in the **dplyr** package. We will use `inner_join()` not `full_join()`.
```{r, message=FALSE,warning=FALSE}
library(dplyr)
```

```{r}
meta <- inner_join(subject, specimen)
head(meta)
```

Let's do one more `inner_join()`
```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q11. How many different isotypes are in the dataset?

```{r}
table(abdata$isotype)
```

> Q. How many antigens?

```{r}
table(abdata$antigen)
```

> Q12. What are the different `$dataset` values in `abdata` and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

Let's focus on the IgG isotype
```{r}
igg <- abdata %>% 
  filter(isotype == "IgG")
head(igg)
```

> Q13. Make a summary boxplot of IgG titer for all antigens.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() 
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigen "PT", "FIM2/3", and "FHA" appear to have the widest range of values. These antigens are found in both the wP and aP vaccines. 

> Q. Is there a difference for these responses between aP and wP individuals?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```
> Q. Is there a difference with time (i.e. before and after bosster shot)?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit) +
  theme_bw()
```




We will plot day relative to boost VS normalized MFI (IgG titer). Filter data to only the 2021 dataset. 

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset", isotype == "IgG", antigen == "PT")

ggplot(abdata.21) + 
  aes(planned_day_relative_to_boost, MFI_normalised, col = infancy_vac, group = subject_id) +
  geom_point(alpha = 0.25) +
  geom_line(alpha = 0.25) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_vline(xintercept = 14, lty = 2) +
  geom_smooth(aes(group = infancy_vac), se = F, span = 0.3) +
  labs(title = "2021 Dataset IgG PT", subtitle = "Dashed lines indicate Day 0 (pre-boost), 14 (apparent peak levels)", x = "Day Relative to Boost", y= "Normalized MFI")
```


```{r}
ggplot(abdata.21) +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

