---
title: "BIMM 143 Class 15"
author: Xaler Lu (A17388454)
format: pdf
toc: true
---

## Background

We will practice with both GitHub and RStudio to streamline workflow. Github is a distributed version control system (D-VCS) that record changes to files over time and track versions over time. It helps make coding collaborative and social, while organizing and archiving files. 

To get started with GitHub, \
1) Create a repository on <Github.com> (remote) \
2) Clone locally using RStudio sync local to remote: pull, stage, commit, push \
3) Explore github.com files, commit history, README \
4) Get exposure to project-oriented workflows in action \

## Using Git

We can use Git from the RStudio "git pane" in the upper right or from the "Terminal Tab".

1) Initiate a git repository by using `git init` in the Terminal Tab \
2) Do your stuff (same as ever) \
3) Select the files that you want to **stage** by using `git add FILENAME` in the Terminal Tab \ 
4) Commit by using `git commit -m "MESSAGE"` in the Terminal Tab \
OPTIONAL) Use `git log` in the Terminal Tab to get a list of changes \

## Connecting RStudio to GitHub

To connect the GitHub website and RStudio \
1) Sign up for a GitHub account. \
2) Use your username and a **Personal Access Token (PAT)** \
3) Use `create_github_token()` from the `usethis` package \
4) Use `gitcreds_set()` from the `gitcreds` package

```{r,message = FALSE, warning= FALSE}
library(usethis)
library(gitcreds)
```

```{r}
create_github_token()
```













